---
title: "Data science final guideline"
author: "Yi Huang, Yuchen Zhang, Shun Xie"
date: "2023-04-28"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 2
  fontsize: 10pt
  geometry: margin=0.5in
  header-includes:
    -\usepackage{fancyhdr}
    -\usepackage{lipsum}
    -\pagestyle{fancy}
    -\fancyhead[R]{\thepage}
    -\fancypagestyle{plain}{\pagestyle{fancy}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      fig.width = 8, 
                      fig.height = 6,
                      out.width = "90%"
                      fig.path = "images/"
)
```
```




```{r, message=FALSE, warning=FALSE}
library(caret)
library(corrplot)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(ISLR)
library(glmnet)
library(mgcv)
library(nlme)
library(earth)
library(vip)
library(pROC)
library(gridExtra)
library(doParallel) # parallel computing
```

## Introduction
```{r}
#load data
load("data/recovery.RData")

#get data from 2 team members uni
set.seed(3554)
dat1 <- dat[sample(1:10000, 2000),]
set.seed(4437)
dat2 <- dat[sample(1:10000, 2000),]

#combine the data and discard duplicates
dat_temp <- rbind(dat1, dat2)
dat <- dat_temp[!duplicated(dat_temp$id),]

#dat is now the final data and it has the length as following:
length(dat$id)

# save(dat, file = "data/covid_recovery.RData")
```


```{r}
# data manipulation
data = 
  dat %>% 
  mutate(study = factor(dat$study),
         gender = factor(dat$gender),
         hypertension = factor(dat$hypertension),
         diabetes = factor(dat$diabetes),
         vaccine = factor(dat$vaccine),
         severity = factor(dat$severity)) %>% 
  dplyr::select(-id)
head(data)

 
#Split data into 70-300, using the third member's uni
set.seed(2337)
indexTrain <- createDataPartition(y = data$recovery_time, p = 0.7, list = FALSE)

# training data
train_data <- data[indexTrain,]
# matrix of predictors
train_x <- model.matrix(recovery_time~.,train_data)[,-1]
# vector of response
train_y <- train_data$recovery_time

# test data
test_data <- data[-indexTrain,]
# matrix of predictors
test_x <- model.matrix(recovery_time~.,test_data)[,-1]
# vector of response
test_y <- test_data$recovery_time
```


# primary analysis

```{r}
# 10-fold cv
ctrl <- trainControl(method = "cv",
                     number = 10)
```

## boosting

```{r}

gbm.grid <- expand.grid(n.trees = c(500,1000,2000,3000,4000,5000),
                        interaction.depth = 1:5,
                        shrinkage = c(0.005,0.01),
                        n.minobsinnode = c(1))

no_cores <- detectCores() - 1
cl <- makePSOCKcluster(no_cores)
registerDoParallel(cl)

set.seed(8)
gbm.fit = train(recovery_time~. ,
                  train_data,
                  tuneGrid = gbm.grid,
                  trControl = ctrl,
                  method = "gbm",
                  verbose = FALSE)

stopCluster(cl)
registerDoSEQ()


ggplot(gbm.fit, highlight = TRUE)

```


```{r}
gbm.fit$bestTune
```


test error
```{r}
gbm.fit.predict = predict(gbm.fit, newdata = test_data)
RMSE(gbm.fit.predict, test_y)
```

```{r}
#vip
summary(gbm.fit$finalModel, las = 2, cBars = 16, cex.names = 0.6)
```


```{r}
#pdp
features = c("bmi", "study", "height", "LDL", "weight", "vaccine")

pdps <- lapply(features, function(x) {
  partial(gbm.fit, pred.var = x, which.class = 2,
          prob = TRUE, plot = TRUE, plot.engine = "ggplot2")
})
grid.arrange(grobs = pdps,  ncol = 3)
```

#class

```{r}
#consider time to recovery as a binary outcome (>30 days vs. <= 30 days) and develop a prediction model for this binary outcome.
train_class = 
  train_data %>% 
  mutate(recovery_time = ifelse(recovery_time>30, "long", "short"))

test_class = 
  test_data %>% 
  mutate(recovery_time = ifelse(recovery_time>30, "long", "short"))

# matrix of predictors
x_train_class <- model.matrix(recovery_time~.,train_class)[,-1]
# vector of response
y_train_class <- train_class$recovery_time


# matrix of predictors
x_test_class <- model.matrix(recovery_time~.,test_class)[,-1]
# vector of response
y_test_class <- test_class$recovery_time


# ctrl for class
ctrl_class <- trainControl(method = "repeatedcv",
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```

# boosting

```{r}

gbmA.grid = expand.grid(n.trees = c(10000,20000,30000,40000),
                         interaction.depth = 1:3,
                         shrinkage = c(0.0003,0.0005,0.001),
                         n.minobsinnode = 1)

no_cores <- detectCores() - 1
cl <- makePSOCKcluster(no_cores)
registerDoParallel(cl)
set.seed(8)
gbmA.fit = train(recovery_time~. ,
                  train_class,
                  tuneGrid = gbmA.grid,
                  trControl = ctrl_class,
                  method = "gbm",
                  distribution = "adaboost",
                  metric = "ROC",
                  verbose = FALSE)
```


```{r}
ggplot(gbmA.fit, highlight = TRUE)
```


```{r}

#train error
train.pred.boosting = predict(gbmA.fit , newdata = train_class)
train.error.boosting = 1 - sum(train_class$recovery_time == train.pred.boosting)/length(train_class$recovery_time)
sprintf("The train error for boosting is %.3f", train.error.boosting)

#test error
test.pred.mars = predict(gbmA.fit, newdata = x_test_class)
test.error.mars = 1 - sum(test_class$recovery_time == test.pred.boosting)/length(test_class$recovery_time)
sprintf("The test error for boosting is %.3f", test.error.mars)
```



# mars
```{r}
no_cores <- detectCores() - 1
cl <- makePSOCKcluster(no_cores)
registerDoParallel(cl)
set.seed(8)
mars.fit = train(x = train_class,
                        y =  y_train_class,
                         method = "earth",
                         tuneGrid = expand.grid(degree = 1:3, 
                                                nprune = 2:20),
                         metric = "ROC",
                         trControl = ctrl_class)
```


```{r}

#train error
train.pred.mars = predict(mars.fit , newdata = train_class)
train.error.mars = 1 - sum(train_class$recovery_time == train.pred.mars)/length(train_class$recovery_time)
sprintf("The train error for MARS is %.3f", train.error.mars)

#test error
test.pred.mars = predict(mars.fit, newdata = test_class)
test.error.mars = 1 - sum(test_class$recovery_time == test.pred.mars)/length(test_class$recovery_time)
sprintf("The test error for MARS is %.3f", test.error.mars)
```

```



